---
title: "datos_nanopore doc"
author: "Ayesha"
date: "2025-01-21"
output: html_document
---
#Basecalling: 
(if needed) used to convert raw FAST5 into FastQ files
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#LongQC / MiniOnQC
(if needed )These are used to process raw raw FAST5 files produced by Oxford Nanopore sequencing


```{r cars}
#if the Rapid Sequencing Kit was used i.e. SQK-NSK007, SQK-MAP006, run command with "ont-rapid" 
#if ligation-based kits are used, run command with "ont-ligation"
#if you are unsure, use porechop (below) that will show the kit while scanning entire data for any known adapters

# Define directories
input_dir="q11L500results"
output_dir="longQC_output"

# Path to LongQC directory
longqc_dir="/home/ayesha/LongQC"

# Create output directory if it doesn't exist
mkdir -p "$output_dir"

# Run LongQC with the correct preset
python3 "$longqc_dir/longQC.py" sampleqc -x ont-ligation -o "$output_dir" "$input_dir"

echo "LongQC analysis complete. Results saved in $output_dir."


```

##Install requirements

```{r}
conda create -n nanopore_env python=3.9 -y
conda activate nanopore_env

 # Install NanoFilt, NanoStat, NanoPlot (python-based tools)
pip install NanoFilt NanoPlot NanoStat

# Install Porechop
pip install porechop 
#if porechop doesnt install, use the following as alternative
pip install cutadapt

#Download and compile Minimap2:
# Download Minimap2
git clone https://github.com/lh3/minimap2.git
cd minimap2
make
# Add minimap2 to PATH (optional)
export PATH=$PATH:$(pwd)
```

##1. Quality Control (FastQC)
assess the quality of the raw sequencing data (FASTQ files).
It generates reports that highlight issues such as:
Adapter contamination
Low base quality
GC content distribution
Overrepresented sequences
After running FastQC, you'll have a good overview of the quality of your data.
```{r}
#increase the space before running fastqc because it takes a lot of storage
export _JAVA_OPTIONS="-Xmx2g"

#go into the folder where fastq files are located, run the code:
 fastqc *.fastq

#if you have to run cfastqc for specific files, change the code snipet below in correspondence to name of fastq files
fastqc barcode{53..74}_filtered_trimmed.fastq
```

##2. Filtering (Nanofilt)
1. Perform quality checks on fastq files (e.g., length distribution, quality scores).
2. Filter out low-quality reads using `NanoFilt` 
```{r}
#for it, enter the folder where your reads folder is located, The for loop iterates through all .fastq files in the reads_extracted folder.
#!/bin/bash

# Create results directory if it doesn't exist
mkdir -p results

# Loop through each FASTQ file in the reads_extracted folder
for file in reads_extracted/*.fastq; do
    # Extract the base name of the file (e.g., sample1 from sample1.fastq)
    base=$(basename "$file" .fastq)

    # Run NanoFilt for each file and save the output in the results directory
    NanoFilt -q 9 -l 1000 "$file" > "results/${base}_filtered.fastq" #(recommended)
    
    NanoFilt -l 100 "$file" > "results/${base}_filtered.fastq"


    echo "Processed $file -> results/${base}_filtered.fastq"
done

#the filtered fastq reads will be saved in "results" folder.
```

##3.Adapter Trimming and Barcode Removal (Porechop)
focused on removing adapter sequences (such as barcodes and sequencing adapters), which should be done after filtering the reads to remove low-quality sequences.

```{r pressure, echo=FALSE}
#!/bin/bash

# Define directory containing the filtered FASTQ files
input_dir="/home/ayesha/v1v9nano/q11L500results"  # Change this if needed
output_dir="/home/ayesha/v1v9nano/q11L500results_trimmed"  # Output directory for trimmed files

# Create output directory if it doesn't exist
mkdir -p "$output_dir"

# Loop through all FASTQ files in the input directory
for file in "$input_dir"/*.fastq; do
    # Extract base name (remove directory and extension)
    base=$(basename "$file" .fastq)
    
    # Run Porechop to trim adapters (Porechop auto-detects the adapters)
    porechop -i "$file" -o "$output_dir/${base}_trimmed.fastq"
    
    # Print out status
    echo "Trimmed $file -> $output_dir/${base}_trimmed.fastq"
done

echo "Porechop trimming completed!"

```

#Multiqc report

```{r}
#enter the folder where files are located
 multiqc .
```

#4. Import data into QIIME2

```{r}
qiime tools import \
  --type 'SampleData[SequencesWithQuality]' \
  --input-path /home/ayesha/v1v9nano/manifest_fastqc_trimmed.tsv \
  --input-format SingleEndFastqManifestPhred33V2 \
  --output-path demux.qza


 qiime demux summarize \
  --i-data demux.qza \
  --o-visualization demux.qzv

#if manifest file is needed for importing:

# Set the directory containing your FASTQ files
FASTQ_DIR="/home/ayesha/v1v9nano/q11L500results_trimmed"

# Create a new manifest file
echo "sample-id,absolute-filepath,direction" > manifest_fastqc_trimmed.csv

# Add each FASTQ file to the manifest file
for file in $FASTQ_DIR/*_filtered_trimmed.fastq; do
    # Get the sample ID from the filename (assuming barcode prefix)
    sample_id=$(basename "$file" _filtered_trimmed.fastq)

    # Add the file path to the manifest
    echo "$sample_id,$file,forward" >> manifest_fastqc_trimmed.csv
done

#if needed, convert the .csv manifest into .tsv file
 sed 's/,/\t/g' /home/ayesha/v1v9nano/manifest_fastqc_trimmed.csv > /home/ayesha/v1v9nano/manifest_fastqc_trimmed.tsv
 
 
```
#5. 
We usually use DADA2 denoise but since our reads lengths are highly variable, we will use VSEARCH OTU Clustering
```{r}
qiime vsearch dereplicate-sequences \
  --i-sequences demux.qza \
  --o-dereplicated-table table.qza \
  --o-dereplicated-sequences rep-seqs.qza

qiime vsearch cluster-features-de-novo \
  --i-sequences rep-seqs.qza \
  --i-table table.qza \
  --p-perc-identity 0.97 \ #sequences with ≥97% similarity will be grouped into the same OTU, a standard threshold for 16S rRNA analyses.
  --o-clustered-table clustered_table.qza \
  --o-clustered-sequences clustered_seqs.qza

 #if the second step takes alot of time, use cpu cores by using thread command as follows:
qiime vsearch cluster-features-de-novo \
  --i-table table.qza \
  --i-sequences rep-seqs.qza \
  --p-perc-identity 0.97 \
  --p-threads 10 \
  --o-clustered-table clustered_table.qza \
  --o-clustered-sequences clustered_seqs.qza
#if you are using server, write --p-threads 6 
```
Train the classifier and visualize taxonomy results
```{r}
qiime feature-classifier classify-sklearn \
  --i-classifier silva-138-full-length-classifier.qza \
  --i-reads clustered_seqs.qza \
  --o-classification taxonomy.qza

qiime metadata tabulate \
  --m-input-file taxonomy.qza \
  --o-visualization taxonomy.qzv

```

Phylogenetic Tree
```{r}
qiime phylogeny align-to-tree-mafft-fasttree \
  --i-sequences clustered_seqs.qza \
  --o-alignment aligned-rep-seqs.qza \
  --o-masked-alignment masked-aligned-rep-seqs.qza \
  --o-tree unrooted-tree.qza \
  --o-rooted-tree rooted-tree.qza

```

